A hidden Active Galactic Nuclei population: the first radio luminosity functions by process

### 2024-08-23 src/tex/ms.tex
[showyourwork] overleaf sync

% Title
\title[TBD]{Direct comparison between simulations and observations: radio emission separated into star formation and AGN activity}

@@

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The extent to which active galactic nuclei (AGN) impact galaxy formation is a major open question in astrophysics. It is clear that there is an interaction, from both observational and theoretical standpoints. Observations have revealed tight scaling relations between the mass of a super-massive black hole and its host galaxy \citep[see, e.g.][and references therein]{kormendy_coevolution_2013}, while cosmological simulations require some form of AGN feedback \citep{bower_breaking_2006,croton_many_2006} to be able to reproduce the observed galaxy population. Galaxies build up their stellar mass through star formation, while super-massive black holes are expected to provide either negative or positive feedback \hl{citations!} to either suppress or stimulate galaxy growth.

@@

\subsection{Simulations}
\label{subsec:simba}
\hl{Nicole}\\
We use the SIMBA cosmological hydrodynamic simulation which accounts for an H$_{2}$ based star formation rate following the sub-grid model of \hl{Krumholtz $\&$ Gnedin (2011)}. Additionally, SIMBA models the growth of supermassive black holes via a two-mode prescription based on the temperature of gas within a distance, $R_{0}$, calculated over the 256 nearest particles limited to 2$h^{-1}$kpc. 
It is commonplace for cosmological simulations to employ the growth of SMBHs via the Bondi rate \hl{citations EAGLE, HORIZON AGN, Illustris, TNG}. In SIMBA, the Bondi rate is employed when gas within $R_{0}$ has temperatures T$>10^{5}$K, referred to as the hot mode. What makes SIMBA unique is the additional contribution to the accretion rate via a torque limited accretion flow when gas has temperatures T$<10^{5}$K, or in the cold mode. 
SIMBA also employs a physically motivated model for AGN feedback in the form of AGN winds \hl{cite Perna}, as well as AGN jets with velocities that increase inversely proportional to the Eddington rate in accordance with \hl{Heckman and Best 2014}.
SIMBA therefore provides a platform to study jet-mode AGN in a way that has not been explored before. 
Specifically, we select galaxies in SIMBA that have full-velocity jets, that is, when the Eddington rate drops below 0.02 (but non-zero) and where the black hole mass is larger than M$_{\rm BH}>10^{8}$M$_{\odot}$. For this population we are able to calculate the amount of radio emission due to star formation processes following \hl{Condon 1992} as well as that from accretion processes following \hl{Kording, Jester and Fender 2008}. Details of this method can be found in \hl{Thomas2021}. We can therefore calculate the contribution to the radio luminosity function from star formation and AGN accretion processes. We can additionally separate the galaxy population into star forming and quiescent galaxies, radio-excess, or radiatively efficient and inefficient, based on the sources black hole and host properties.
\section{Methods}

@@

To illustrate the differences in RLFs constructed by activity vs. by galaxy classification we divide the activity RLF by the galaxy RLF and show the ratio in the bottom right hand panel in Fig.~\ref{fig:rlfs}, for both the AGN and SF category. We see that above 10$^{24}\,$\wphz\ the ratio for AGN is unity, which is expected where radio excess sources dominate the population. Below that, the ratio for AGN climbs up to $\sim$1.25 and finally almost up to 2 at the lowest radio powers probed. The ratio for SF is always below unity, but it does not deviate as much as the ratio for AGN. This is because the SF population has larger source counts, so the total impact of moving some flux density from the SF to AGN population is less extreme.

@@

In Fig.~\ref{fig:evolution} we show a comparison between the observed RLFs and those generated from SIMBA (see \S~\ref{subsec:simba}), and the redshift evolution for both. The left set of panels shows the results for SF and the right set of panels shows the results for AGN. \hl{We cut off the RLFs below a flux limit where they are not constrained}. We use the same redshift bins across all samples for consistency. We reproduce previous observational results that while the SF population has a strong redshift evolution, the AGN population does not. For the simulations, the behaviour is qualitatively similar, although the AGN population does have mild redshift evolution whereas the observations do not.

The bottom panels in Fig~\ref{fig:evolution} show the ratio of activity to galaxy RLFs, the same as in the bottom right hand panel of Fig~\ref{fig:rlfs}. Comparing these ratios between observations and simulations, we see qualitatively the same behaviour, although the simulations reach much lower and higher values for both the SF and AGN populations, respectively. In the lowest redshift bin, the ratio for the simulation stabilises to around unity for \llof $<10^{23}\,$\wphz , and decreases rapidly above this, reaching much smaller values than the observations. Higher redshift bins probe successively higher luminosity ranges, also reaching small values for the ratio between activity and galaxies. This is consistent with the fact that at higher redshifts, most radio AGN in SIMBA are hosted in SFGs, and thus the \textit{galaxy} RLF for SIMBA SFGs is inflated with AGN emission.

For the AGN population (Fig~\ref{fig:evolution}, right set of panels), the ratio between observed activity and galaxy RLFs  converges to unity above $\sim$\llof $10^{24}\,$\wphz\ for the lowest redshift bin. This convergence happens at higher luminosity values for higher redshifts. Below these luminosities, across all redshift bins, the ratio increases. The maximum value for each redshift bin approaches $\sim$2, although the fact that in the lowest redshift bin the value stabilises at $\sim1.3$ and only climbs higher at the very lowest luminosities (\llof $<10^{22}\,$\wphz ) may be due to incompleteness. If we disregard the lowest luminosity bins, it seems that we see a 30 - 50 percent increase in AGN activity over what is measured when using galaxy classifications alone, for all redshifts. The ratio for the SIMBA RLFs shows strong redshift evolution, and much higher ratio values. The RLFs themselves turn over, whereas the observed RLFs do not. This is likely a consequence of the fact that the AGN emission in SIMBA is due to high-velocity jets, whereas there is increasing observational evidence for low-luminosity jets or AGN emission \citep{macfarlane_radio_2021,yue_novel_2024}. The high values reached by the ratio of RLFs again is expected as many high-$z$ radio AGN in SIMBA are hosted in SFGs. Shifting this AGN emission to AGN \textit{activity} when it is counted in the SFG \textit{galaxy} population generates high values for the activity to galaxy ratio.

>>>>

% Title
\title[TBD]{A hidden Active Galactic Nuclei population: the first radio luminosity functions by process}

@@

% Main body with filler text
\section{Introduction}
\label{sec:intro}
The extent to which active galactic nuclei (AGN) impact galaxy formation is a major open question in astrophysics. It is clear that there is an interaction, from both observational and theoretical standpoints. Observations have revealed tight scaling relations between the mass of a super-massive black hole and its host galaxy \citep[see, e.g.][and references therein]{kormendy_coevolution_2013}, while cosmological simulations require some form of AGN feedback \citep{bower_breaking_2006,croton_many_2006} to be able to reproduce the observed galaxy population. Galaxies build up their stellar mass through star formation, while super-massive black holes are expected to provide either negative or positive feedback to either suppress or stimulate galaxy growth, although whether we can definitively pinpoint which type of feedback is an open question \citep{ward_cosmological_2022}.

@@

\subsection{Simulations}
\label{subsec:simba}

We use the SIMBA cosmological hydrodynamic simulation which accounts for an H$_{2}$ based star formation rate following the sub-grid model of \cite{krumholz_comparison_2011}. Additionally, SIMBA models the growth of supermassive black holes via a two-mode prescription based on the temperature of gas within a distance, $R_{0}$, calculated over the 256 nearest particles limited to 2$h^{-1}$kpc. 
It is commonplace for cosmological simulations to employ the growth of SMBHs via the Bondi rate \hl{citations EAGLE, HORIZON AGN, Illustris, TNG}. In SIMBA, the Bondi rate is employed when gas within $R_{0}$ has temperatures T$>10^{5}$K, referred to as the hot mode. What makes SIMBA unique is the additional contribution to the accretion rate via a torque limited accretion flow when gas has temperatures T$<10^{5}$K, or in the cold mode. 
SIMBA also employs a physically motivated model for AGN feedback in the form of AGN winds \citep{perna_x-raysdss_2017}, as well as AGN jets with velocities that increase inversely proportional to the Eddington rate in accordance with \cite{heckman_coevolution_2014}.
SIMBA therefore provides a platform to study jet-mode AGN in a way that has not been explored before. 
Specifically, we select galaxies in SIMBA that have full-velocity jets, that is, when the Eddington rate drops below 0.02 (but non-zero) and where the black hole mass is larger than M$_{\rm BH}>10^{8}$M$_{\odot}$. For this population we are able to calculate the amount of radio emission due to star formation processes following \cite{condon_radio_1992} as well as that from accretion processes following \cite{kording_measuring_2008}. Details of this method can be found in \cite{thomas_radio_2021}. We can therefore calculate the contribution to the radio luminosity function from star formation and AGN accretion processes. We can additionally separate the galaxy population into star forming and quiescent galaxies, radio-excess, or radiatively efficient and inefficient, based on the sources black hole and host properties.

\section{Methods}

@@

To illustrate the differences in RLFs constructed by activity vs. by galaxy classification we divide the activity RLF by the galaxy RLF for the observations and show the ratio in the bottom right hand panel in Fig.~\ref{fig:rlfs}, for both the AGN and SF category. We see that above 10$^{24}\,$\wphz\ the ratio for AGN is unity, which is expected where radio excess sources dominate the population. Below that, the ratio for AGN climbs up to $\sim$1.25 and finally almost up to 2 at the lowest radio powers probed. The ratio for SF is always below unity, but it does not deviate as much as the ratio for AGN. This is because the SF population has larger source counts, so the total impact of moving some flux density from the SF to AGN population is less extreme.

@@

In Fig.~\ref{fig:evolution} we show a comparison between the observed RLFs and those generated from SIMBA (see \S~\ref{subsec:simba}), and the redshift evolution for both.  We use similar redshift bins to \cite{kondapally_cosmic_2022} and \cite{cochrane_lofar_2023} for consistency. The left set of panels shows the results for SF and the right set of panels shows the results for AGN. \hl{We cut off the RLFs below a flux limit where they are not constrained}. We use the same redshift bins across all samples for consistency. We reproduce previous observational results that while the SF population has a strong redshift evolution, the AGN population does not. For the simulations, the behaviour is qualitatively similar, although the AGN population does have mild redshift evolution whereas the observations do not.

The bottom panels in Fig~\ref{fig:evolution} show the ratio of activity to galaxy RLFs, the same as in the bottom right hand panel of Fig~\ref{fig:rlfs} (although note that the y-scale is logarithmic for the AGN). Comparing these ratios between observations and simulations, we see qualitatively the same behaviour, although the simulations reach much lower and higher values for both the SF and AGN populations, respectively. Looking first at the SF population (left set of panels), in the lowest redshift bin the ratio for the simulation stabilises to around unity for \llof $<10^{23}\,$\wphz , and decreases rapidly above this, reaching much smaller values than the observations. Higher redshift bins probe successively higher luminosity ranges, also reaching small values for the ratio between activity and galaxies. This is consistent with the fact that at higher redshifts, most radio AGN in SIMBA are hosted in SFGs, and thus the \textit{galaxy} RLF for SIMBA SFGs is inflated with AGN emission; this makes the ratio of SF activity to galaxy approach zero.

For the AGN population (Fig~\ref{fig:evolution}, right set of panels), the ratio between observed activity and galaxy RLFs converges to unity above $\sim$\llof $10^{24}\,$\wphz\ for the lowest redshift bin. This is driven by the fact that the flux density of radio excess sources is entirely attributed to AGN, and these sources tend to have higher radio luminosities. The convergence happens at higher luminosity values for higher redshifts. Below these luminosities, across all redshift bins, the ratio increases. The maximum value for each redshift bin approaches $\sim$2, although the fact that in the lowest redshift bin the value stabilises at $\sim1.3$ and only climbs higher at the very lowest luminosities (\llof $<10^{22}\,$\wphz ) may be due to incompleteness. If we disregard the lowest luminosity bins, it seems that we see approximately a 30 - 50 percent increase in AGN activity over what is measured when using galaxy classifications alone, for all redshifts. The ratio for the SIMBA RLFs shows strong redshift evolution, and much higher ratio values. The RLFs themselves turn over, whereas the observed RLFs do not. This is likely a consequence of the fact that the AGN emission in SIMBA is due to high-velocity jets, whereas there is increasing observational evidence for low-luminosity jets or AGN emission \citep{macfarlane_radio_2021,yue_novel_2024}. The high values reached by the ratio of RLFs again is expected as many high-$z$ radio AGN in SIMBA are hosted in SFGs. Shifting this AGN emission to AGN \textit{activity} when it is counted in the SFG \textit{galaxy} population generates high values for the activity to galaxy ratio.
